---
output: github_document
bibliography: bibliography.bibtex
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
library(tidyverse)
library(cowplot)
```

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# PbIso R package

<!-- badges: start -->
<!-- badges: end -->

The package `PbIso` is a free and open R toolbox for commonly used calculations and plots of Pb-Pb isotope data.

## Installation

The development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("shereearmistead/PbIso")
library(PbIso)
```

```{r, echo = FALSE}
library(PbIso)
```
#	Functions {#calcs}

All of the functions in `PbIso` take one or more of the basic input parameters t (time (Ma)), x (^206^Pb/^204^Pb), y (^207^Pb/^204^Pb) and z (^208^Pb/^204^Pb) to perform the calculations. For advanced usage, the functions can also optionally take the values for different model parameters given in Table 1. See documentation for details on what input parameters are required or allowed in each function. The calculations and functions used in `PbIso` assume a starting composition and model following @RN10 2nd stage model. These values can be overridden in R if an alternative model (e.g. @RN10 1st stage or others) is preferred. 


```{r table1, echo = FALSE, fig.pos='H'}
model.params <- read.csv("ModelParams.csv")

knitr::kable(model.params, justify = c("left", "centre"), caption = "Default model parameters used in `PbIso` functions. These can be manually changed if an alternative model is preferred.", "pipe", escape = FALSE)
```

## The evolution of radiogenic Pb isotopes with time {#basiccalcs}

Calculate the ^206^Pb/^204^Pb, ^207^Pb/^204^Pb and ^208^Pb/^204^Pb ratios at a given time (assuming Stacey & Kramers (1975) 2nd stage model parameters), for example at 2700 Ma:

```{r example}
Calc64(2700)
Calc74(2700)
Calc84(2700)
```

```{r fig1, echo = FALSE, fig.cap = "Pb isotopic evolution through time, a) evolution of ^206^Pb/^204^Pb and ^207^Pb/^204^Pb; b) evolution of ^206^Pb/^204^Pb and ^208^Pb/^204^Pb. Isochrons for 3000 Ma, 2000 Ma, 1000 Ma and 0 Ma shown as dashed lines in both plots.", dpi=1000, fig.width = 10, fig.height= 4}
model <- modelcurve(0:3700)
model.subset <- model[seq(1, nrow(model), 500),]
X1 <- 11.152            # 206Pb/204Pb at 3700 Ma - Stacey & Kramers 1975
Y1 <- 12.998            # 207Pb/204Pb at 3700 Ma - Stacey & Kramers 1975
Z1 <- 31.230            # 208Pb/204Pb at 3700 Ma - Stacey & Kramers 1975
T1 <- 3700       # Age 3700 Ma
p1 <- ggplot(model, aes(x = x, y = y)) +
    labs(y = expression(paste(""^207, "Pb/", ""^204, "Pb")), 
       x = expression(paste(""^206, "Pb/", ""^204, "Pb"))) +
  geom_line() +
  geom_point(x = X1, y = Y1) +
  annotate('text', x = X1, y = Y1, label = "T1 = 3700 Ma", hjust=-0.1, vjust=0.75) +
  geom_point(data = model.subset) +
  geom_text(data = model.subset, aes(label = t), hjust=1, vjust=-0.5) +
  geom_abline(slope = geochron76slope(3000), intercept = geochron76yint(3000), linetype = 2) +
  geom_abline(slope = geochron76slope(2000), intercept = geochron76yint(2000), linetype = 2) +
  geom_abline(slope = geochron76slope(1000), intercept = geochron76yint(1000), linetype = 2) +
  geom_abline(slope = geochron76slope(0), intercept = geochron76yint(0), linetype = 2) +
  theme_bw(base_size = 18) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  ylim(12.75, 16)

p2 <- ggplot(model, aes(x = x, y = z)) +
    labs(y = expression(paste(""^208, "Pb/", ""^204, "Pb")), 
       x = expression(paste(""^206, "Pb/", ""^204, "Pb"))) +
    geom_line() +
    geom_abline(slope = geochron86slope(0), intercept = geochron86yint(0), linetype = 2) +
    geom_abline(slope = geochron86slope(1000), intercept = geochron86yint(1000), linetype = 2) +
    geom_abline(slope = geochron86slope(2000), intercept = geochron86yint(2000), linetype = 2) +
    geom_abline(slope = geochron86slope(3000), intercept = geochron86yint(3000), linetype = 2) +
    geom_point(data = model.subset) +
    geom_point(x = X1, y = Z1) +
    annotate('text', x = X1, y = Z1, label = "T1 = 3700 Ma", hjust=-0.1, vjust=0.75) +
    geom_text(data = model.subset, aes(label = t), hjust=1, vjust=-1) +
    theme_bw(base_size = 18) +
    theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
    scale_y_continuous(limits = c(31,39), breaks = c(32, 34, 36, 38))
plot_grid(p1, p2, labels = 'auto')
```
## Model age

To calculate the model age of a hypothetical sample with ^206^Pb/^204^Pb = 13.5 and ^207^Pb/^204^Pb = 14.5: 
```{r display CalcModAge, warning=FALSE}
CalcModAge(13.5, 14.5)
```

##	Model source $\mu$ (^238^U/^204^Pb)

To calculate the model source $\mu$ (^238^U/^204^Pb), we need to know the age of the sample. If we have a hypothetical sample with known age of t = 2700 Ma; ^206^Pb/^204^Pb = 13.5, and ^207^Pb/^204^Pb = 14.5, and accepting the default model values, we can calculate the model source $\mu$ by:

```{r display CalcMu}
CalcMu(2700, 13.5, 14.5)
```

##	Time-integrated $\kappa$ (^232^Th/^238^U) {#kappa}

To calculate the time-integrated $\kappa$ (^232^Th/^238^U), let's assume the same sample with ^206^Pb/^204^Pb = 13.5; t = 2700 Ma and now with ^208^Pb/^204^Pb = 33:

```{r display CalcKa}
CalcKa(2700, 33, 13.5)
```

## Initial Pb isotope ratios {#initial}
To calculate initial Pb isotope ratios:

```{r}
Calc64in(2700, 13.5, 14.5)
Calc74in(2700, 13.5, 14.5)
Calc84in(2700, 13.5, 14.5, 33)
```

## Model curves {#modelcurves}

The `modelcurve()` function can be used to produce model curves over different time ranges with optional arguments (see `PbIso` documentation for more information). For example, in the code below, 'mc1' is the Stacey & Kramers (1975) 1st stage model from 4570 Ma to 3700 Ma, 'm2' is the Stacey & Kramers (1975) 2nd stage model from 3700 Ma to present, and 'mc3' is a hypothetical model using Stacey & Kramers (1975) 2nd stage starting composition but with a $\mu_1$ of 8. These three curves are shown in the figure below.

```{r}
mc1 <- modelcurve(3700:4570, T1 = 4570, X1 = 9.307, Y1 = 10.294, Z1 = 29.487, W1 = 33.21, Mu1 = 7.19)
mc2 <- modelcurve(0:3700)
mc3 <- modelcurve(0:3700, Mu1 = 8)
```

```{r figmc, fig.pos='H', echo = FALSE, fig.cap = "Model curves generated using the `modelcurve()` function for [@RN10] 1st stage (mc1), [@RN10] 2nd stage (mc2) and a hypothetical sample with Mu=8 (mc3)", dpi=1000, fig.width = 10, fig.height= 4}

a <- ggplot(mc1, aes(x = x, y = y)) +
  labs(color='model curve', y = expression(paste(""^207, "Pb/", ""^204, "Pb")), x = expression(paste(""^206, "Pb/", ""^204, "Pb"))) +
  geom_line(aes(color = 'mc1')) +
  geom_line(data = mc2, aes(color = 'mc2')) +
  geom_line(data = mc3, aes(color = 'mc3')) +
  scale_color_manual(values = c('red', 'black', 'blue')) +
  theme_bw(base_size = 16) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), legend.position = c(0.7,0.3)) 

b <- ggplot(mc1, aes(x = x, y = z)) +
  labs(y = expression(paste(""^208, "Pb/", ""^204, "Pb")), x = expression(paste(""^206, "Pb/", ""^204, "Pb"))) +
  geom_line(aes(color = 'mc1')) +
  geom_line(data = mc2, aes(colour = 'mc2')) +
  geom_line(data = mc3, aes(colour = 'mc3')) +
  scale_color_manual(values = c('red', 'black', 'blue')) +
  theme_bw(base_size = 16) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), legend.position='none')

plot_grid(a, b)
```

## Geochron lines

To generate geochron lines for a given time (t), the slope and y-intercept are needed. To calculate the slope of a geochron line on a ^206^Pb/^204^Pb vs. ^207^Pb/^204^Pb plot, the function `geochron76slope()` is used, which takes the argument t as well as optional arguments (see documentation). The associated y-intercept for that geochron is given by the function `geochron76yint()`. These can then be used to plot the geochron line along with a model curve. Similarly, to calculate the geochron slope and y-intercept on a ^206^Pb/^204^Pb vs. ^208^Pb/^204^Pb plot, the functions `geochron86slope()` and `geochron86yint()` can be used. These values can then be used to plot geochron lines, or the function can be called directly, for example by using `abline(a = geochron76yint(2700), b = geochron76slope(2700))` in base R plotting, or `geom_abline(slope = geochron76slope(2700), intercept = geochron76yint(2700))` in ggplot. To use this function to generate the slope and y-intercept of a geochron at time 2700 Ma:

```{r}
geochron76slope(2700)
geochron76yint(2700)
```

```{r}
geochron86slope(2700)
geochron86yint(2700)
```

# References
